---
title: "Reproducible Research Project 1"
output: html_document
---

##Loading and preprocessing the data

Install Packages
```{r}
library(ggplot2)
library(dplyr)
```

Load the data
```{r}
activity <- read.csv("activity.csv")
```
Convert date column to datetime
```{r}
activity$date <- as.Date(activity$date)
```

##What is mean total number of steps taken per day?

Plot a histogram of the total number of steps taken each day
```{r}
ggplot(na.omit(activity), aes(x=date, y=steps)) + geom_bar(stat="identity") + 
  labs(main="Total Steps Per Day", x="Date", y="Total Number of Steps")
```

Calculate the total steps per day
```{r}
totalSteps <- activity %>% group_by(date) %>% filter(!is.na(steps)) %>%
                summarize(totalStepsPerDay = sum(steps, na.rm=TRUE))
totalSteps
```

Calculate the mean steps per day
```{r}
meanSteps<- mean(totalSteps$totalStepsPerDay, na.rm=TRUE)
meanSteps
```

Calculate the median steps per day
```{r}
medianSteps<- median(totalSteps$totalStepsPerDay, na.rm=TRUE)
medianSteps
```

##What is the average daily activity pattern?

Average total steps taken over 5 minute intervals
```{r}
totalIntSteps <- activity %>% group_by(interval) %>% filter(!is.na(steps)) %>%
                summarize(totalStepsPerInt = mean(steps, na.rm=TRUE))
totalIntSteps
```

Plot average total steps taken over 5-minute intervals
```{r}
ggplot(na.omit(totalIntSteps), aes(x=interval, y=totalStepsPerInt)) + geom_line() + 
  labs(main="Average Total Steps Over 5 Minute Intervals", x="Intervals", y="Total Number of Steps")
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
totalIntSteps[which.max(totalIntSteps$totalStepsPerInt),]
```

##Imputing missing values

Total number of missing values in the dataset
```{r}
sum(is.na(activity$steps))
```

Fill in the missing values in the dataset with the mean for that 5-minute interval
```{r}
activity2 <- activity

nas <- is.na(activity2$steps)
intervalMean <- tapply(activity2$steps, activity2$interval, mean, na.rm=TRUE, simplify=TRUE)
activity2$steps[nas] <- intervalMean[as.character(activity2$interval[nas])]
```

Double-check there are no missing values in the dataset
```{r}
sum(is.na(activity2$steps))
```

Calculate number of steps taken per day
```{r}
stepsFilledNAs <- activity2 %>% group_by(date) %>% filter(!is.na(steps)) %>% summarize(steps = sum(steps)) %>% print
```

Plot a histogram of the total number of steps taken each day with NAs filled
```{r}
ggplot(na.omit(stepsFilledNAs), aes(x=date, y=steps)) + geom_bar(stat="identity") + 
  labs(main="Total Steps Per Day", x="Date", y="Total Number of Steps")
```

Calculate the mean steps per day
```{r}
meanSteps2 <- mean(stepsFilledNAs$steps, na.rm=TRUE)
meanSteps2
```

Calculate the median steps per day
```{r}
medianSteps2 <- median(stepsFilledNAs$steps, na.rm=TRUE)
medianSteps2
```

##Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r}
activity2 <- mutate(activity2,weekType = ifelse(weekdays(activity2$date) == "Saturday" | weekdays(activity2$date) == "Sunday", "weekend", "weekday"))

activity2$weekType <- as.factor(activity2$weekType)
head(activity2)
```

Panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days.
```{r}
intervalWeekType <- activity2 %>% group_by(interval, weekType) %>% summarize(steps = mean(steps))

weekTypePlot <- ggplot(intervalWeekType, aes(x=interval, y=steps, color = weekType)) + geom_line() + 
        facet_wrap(~weekType, ncol = 1, nrow=2)
print(weekTypePlot)
```